SHARP II: Dark Matter Substructure, Gravitational Lens Flux Ratio Anomalies, and the Case of B1555+375: Beware the Effect of Edge-on Disks

### 2015-04-25 papers/B1555_EdgeOn/B1555_letter.tex
spell-check of current draft

A common feature of simulations of structure formation is the presence
of thousands of subhaloes that are associated with larger mass haloes \citep[e.g.][]{Springel08}.
However, observations of the Local Group find many fewer satellite
galaxies than are predicted by the simulations, even when taking into
account completeness corrections arising from the limited sky coverage
imposed by the Galactic Plane and the latest discovery of new dwarf galaxies \citep{DES15,Kop15}.  
This is the famous ``missing satellite'' problem \citep{Klypin1999, Moore1999, S07}. In order to
understand this discrepancy, it is imperative to build up a large
sample of satellites/subhaloes in galaxies outside the Local Group.
 However, this process is challenging due to the faintness of the satellite
galaxies.  Furthermore, lower-mass satellites may not be able to
retain the gas needed for ongoing star formation \citep[e.g.,][]{P11},
rendering them effectively dark.  These factors make gravitational
lensing a powerful and promising tool for the detection of
substructure.

@@

While future large-scale surveys, such as the LSST, are expected to
deliver new large samples of gravitationally lensed quasars, in the
near future the discovery of new systems using narrow-line quasar
emission \citep{N14} will provide new invaluable statistical
perspective on current constraints based on flux ratio anomalies.  In
this letter, we explore the idea that not all flux ratio anomalies are
related to the presence of substructure, but that some originate in a
more complex lens galaxy mass distribution than initially
considered. To this end we use new infra-red data from the Strong
lensing at High Angular Resolution Program 
\citep[SHARP;][Fassnacht et al., in prep]{CLASS1}.
The main goal of the SHARP project is to re-observe known
quadruple and Einstein ring lens systems at high angular
resolution by making use of observations with (1) adaptive optics (AO)
on the Keck telescopes, (2) the Hubble Space Telescope, and (3) large
interferometric arrays \citep{SHARP12,V12}. In particular, to better
understand the causes of flux ratio anomalies in gravitational lenses
we focus on the B1555+375 lens system\citep{Marlow99}.  This system was
discovered in the Cosmic Lens All-Sky Survey \citep{CLASS1,CLASS2} and
has a pair of merging images that show a strong flux ratio anomaly at
radio wavelengths.  Follow-up observations ruled out radio propagation
effects as the cause of the anomaly \citep{K03,KD04}.  The B1555+375
system was included in both the observational and theoretical analysis
of \citet{Dalal2002} and \citet{Xu14} and, thus, understanding its
mass distribution is of key importance.  Our analysis makes use of
the new information obtained from the latest SHARP infrared imaging.
The paper is organized as follows: the  SHARP data are
introduced in section 2; in section 3 we present the lens model while
the results of this analysis and their implications are discussed in
section 4.

@@

\section{Lens Modelling}

@@

As a first trial of modelling, we create a single SIE model to check
the performance of a simple lensing potential. Consistent with
previous studies \citep{Marlow99, Xu14}, our single elliptical
potential model cannot predict the flux anomaly in B1555+375.  We then
create two separate models of the lensing potential that both
incorporate the disk.  In the first we treat the lens as a SIE bulge
and an exponential edge-on disk (SIE+expdisk).  In the second model
the disk is represented as a very elongated SIE component (2-SIE).
The lens system is red, with a strong jump in flux between the F814W
(roughly I) and F160W (roughly H) bands (Fig.~\ref{fig:multiband}), so
we assign a lens redshift of $z_\ell = 1.0$ and a source redshift of
$z_s =1.5$. Our steps to explore the parameter space are: (1) Apply
strong priors from SHARP AO image to the edge-on disk position angle
and ellipticity, and use the best-fit model of \citet{Marlow99} as the
starting point for the bulge component and the source position. (2)
Let all parameters without priors vary to reach a temporary best
fit. (3) Start with the best fit in step 2 and let all parameters with
priors vary to reach another best fit (also ensure this best fit is
not far from the priors). (4) Let all parameters vary to reach the
final optimal model.  The full best-fit parameters are shown in Table
2. Both of these models produce lensed images with positions and
fluxes that are good matches to the radio observations.  While both of
the two component models produce significantly improved matches to the
fluxes than the single SIE model, the 2-SIE version has a somewhat
better performance than the SIE+expdisk model in reproducing flux
ratios. The comparison of the observations and our lens modelling is
given in Figure 3 and Tables 3 \& 4.  We have not included an external
shear component into our model because the constraints provided by the
image positions and fluxes would not be sufficient to constrain such a
model. However, we note that the image not affected by dust (image C)
in the Keck AO data is consistent with negligible shear (i.e. shear
strength $\Gamma_s=0.002$), when modelled with the pixelized technique
of \citet{V09}.

@@

%\end{minipage}
\medskip
Comparision between radio data and model-predicted flux ratios. The radio observation data is quoted from Tables 2 and Table 3 in \citet{Marlow99}. 1-SIE model is quoted from Tabel 5 in \citet{Marlow99}.

@@

In this letter, we focus on the effect of more complicated mass
distributions for the lens galaxy on flux ratio anomalies.
Traditionally, flux anomalous lensed quasars are radio loud systems
that are firstly discovered in radio observations. While radio data
provide reliable information on the positions and the fluxes of the
lensed images, they are often unable to detect the lens
galaxies. Moreover, lens modelling based only on the positions of the
four images can be highly degenerate and prior assumptions play a key
role in the interpretation of the observed flux ratio anomaly.  Since
lens galaxies are typically at redshifts of $z \leq 1$, optical and
near-infrared observations are ideal for measuring the light distribution
of lens galaxies and obtaining useful prior information on the underlying
mass distributions. Indeed \citet{SHARP12} showned that high
resolution IR imaging can help to significantly improve the precision
of the lens modelling. This improvement is even more dramatic in the
case of B1555+375, where HST and SHARP AO high resolution imaging has
allowed us to detect, for the fist time, the presence of an edge-on
disk. This new discovered edge-on disk provides a strong prior for our
lens modelling.  With a bulge plus an edge-on disk, our models produce
matches to both the radio image positions and their flux ratio, while
models with only one SIE component \citep{Marlow99,Xu14} fail to
reproduce the fluxes correctly. Our successful modelling indicates
that the kiloparsec-scale structure of the lens galaxy itself is
sufficient to explain the flux ratio anomalies in B1555+375, without
any need to invoke the presence of substructure.

From a theoretical perspective, \citet{Xu14} also find that substructures
cannot explain the observed radio flux anomalies in all the lenses
that were considered. After adding subhaloes from numerical simulations
to these systems, only one out of the eight radio loud systems
achieves the level of the flux anomalies seen in the observations.  The rest of
them are unlikely (only $1 \% - 4 \%$ probability) to have flux
anomalies produced only by substructure, including our target B1555+375.
In fact, among those eight radio lenses, B1555+375 is one of the most
anomalous systems, but it is also one of the most challenging systems
to explain by substructures. \citet{Xu14} have concluded that the failure in
predicting flux ratios is most likely due to oversimplified/improper lens
modelling. They also point out that the SIE component and the external
shear in their model are orthogonal, indicating a possible missing
ingredient in the lens model. Their results are consistent with the SHARP
$K^{\prime}$ band observations, in which an edge-on disk crosses the
radio merging double and which fully accounts for the observed flux ratio anomaly.  In general, it is well known 
\citep[e.g,][]{Ka91} that lens models based only on the locations of lensed
quasar/AGN images have large degeneracies. Therefore, before
concluding that the flux ratios in any given lens system are a result
of substructure, it is critical to explore a range of smooth models
that are informed by additional observations or to do a comprehensive
analysis such as that conducted by \citet{Xu14}. Otherwise, derived
substructure fractions should be considered as upper limits.

@@

%. Modelling with only four image positions from radio data can also cause a lot of degeneracy. With the first system analyzing and finding an edge-on disk originated flux ratio anomaly, our follow up will be explore other similar SHARP systems. In Figure 3, all of these systems have a recognizable edge-on disk in infrared imaging. An edge-on disk structure may also be able to explain the flux ratio anomalies of these systems.
 %From the successful modelling of B1555+375,  we expect more flux anomalous systems requiring complex lens models (e.g. include an exponential disk), rather than just assuming smooth potential. Although the number of radio loud flux anomalous lensed quasars remains small, a new technique using narrow-line emissions will enlarge the sample and bring statistical perspective on substructures in the near future \citep{N14}. If the non-substructure effect from edge-on disks does not included, expected ``false'' substructure detections can serverly bias the data analysis. 
% This breaks the idea that in strongly lensed quasars, radio flux ratio
% anomalies are most likely originated from substructure
% perturbations. Our discovery shows a complex lens galaxy mass
% distribution can dominate flux ratio anomalous phenomena in a strongly
% lensed system.
% We would like to draw attention to the possible impacts on
% substructure searching from this enge-on disk orginated flux
% anomalies. It is a sign that researchers should be careful about the
% flux anomalies fraction in lensed quasars. It may be worth considering
% an upper limit of flux anomalies fraciton from observation results to
% avoid over estimate the number of substructure detections. For further
% analysis on substurcture population, mass function etc., it is always
% important to rule out other non-substructure effects.

>>>>

A common feature of simulations of structure formation is the presence
of thousands of subhalos that are associated with larger mass aloes \citep[e.g.][]{Springel08}.
However, observations of the Local Group find many fewer satellite
galaxies than are predicted by the simulations, even when taking into
account completeness corrections arising from the limited sky coverage
imposed by the Galactic Plane and the latest discovery of new dwarf galaxies \citep{DES15,Kop15}.  
This is the famous ``missing satellite'' problem \citep{Klypin1999, Moore1999, S07}. In order to
understand this discrepancy, it is imperative to build up a large
sample of satellites/subhalos in galaxies outside the Local Group.
 However, this process is challenging due to the faintness of the satellite
galaxies.  Furthermore, lower-mass satellites may not be able to
retain the gas needed for ongoing star formation \citep[e.g.,][]{P11},
rendering them effectively dark.  These factors make gravitational
lensing a powerful and promising tool for the detection of
substructure.

@@

While future large-scale surveys, such as the LSST, are expected to
deliver new large samples of gravitationally lensed quasars, in the
near future the discovery of new systems using narrow-line quasar
emission \citep{N14} will provide new invaluable statistical
perspective on current constraints based on flux ratio anomalies.  In
this letter, we explore the idea that not all flux ratio anomalies are
related to the presence of substructure, but that some originate in a
more complex lens galaxy mass distribution than initially
considered. To this end we use new infra-red data from the Strong
lensing at High Angular Resolution Program 
\citep[SHARP;][Fassnacht et al., in prep]{SHARP12}.
The main goal of the SHARP project is to re-observe known
quadruple and Einstein ring lens systems at high angular
resolution by making use of observations with (1) adaptive optics (AO)
on the Keck telescopes, (2) the Hubble Space Telescope, and (3) large
interferometric arrays \citep{SHARP12,V12}. In particular, to better
understand the causes of flux ratio anomalies in gravitational lenses
we focus on the B1555+375 lens system\citep{Marlow99}.  This system was
discovered in the Cosmic Lens All-Sky Survey \citep{CLASS1,CLASS2} and
has a pair of merging images that show a strong flux ratio anomaly at
radio wavelengths.  Follow-up observations ruled out radio propagation
effects as the cause of the anomaly \citep{K03,KD04}.  The B1555+375
system was included in both the observational and theoretical analysis
of \citet{Dalal2002} and \citet{Xu14} and, thus, understanding its
mass distribution is of key importance.  Our analysis makes use of
the new information obtained from the latest SHARP infrared imaging.
The paper is organized as follows: the  SHARP data are
introduced in section 2; in section 3 we present the lens model while
the results of this analysis and their implications are discussed in
section 4.

@@

\section{Lens Modeling}

@@

As a first trial of modeling, we create a single SIE model to check
the performance of a simple lensing potential. Consistent with
previous studies \citep{Marlow99, Xu14}, our single elliptical
potential model cannot predict the flux anomaly in B1555+375.  We then
create two separate models of the lensing potential that both
incorporate the disk.  In the first we treat the lens as a SIE bulge
and an exponential edge-on disk (SIE+expdisk).  In the second model
the disk is represented as a very elongated SIE component (2-SIE).
The lens system is red, with a strong jump in flux between the F814W
(roughly I) and F160W (roughly H) bands (Fig.~\ref{fig:multiband}), so
we assign a lens redshift of $z_\ell = 1.0$ and a source redshift of
$z_s =1.5$. Our steps to explore the parameter space are: (1) Apply
strong priors from SHARP AO image to the edge-on disk position angle
and ellipticity, and use the best-fit model of \citet{Marlow99} as the
starting point for the bulge component and the source position. (2)
Let all parameters without priors vary to reach a temporary best
fit. (3) Start with the best fit in step 2 and let all parameters with
priors vary to reach another best fit (also ensure this best fit is
not far from the priors). (4) Let all parameters vary to reach the
final optimal model.  The full best-fit parameters are shown in Table
2. Both of these models produce lensed images with positions and
fluxes that are good matches to the radio observations.  While both of
the two component models produce significantly improved matches to the
fluxes than the single SIE model, the 2-SIE version has a somewhat
better performance than the SIE+expdisk model in reproducing flux
ratios. The comparison of the observations and our lens modeling is
given in Figure 3 and Tables 3 \& 4.  We have not included an external
shear component into our model because the constraints provided by the
image positions and fluxes would not be sufficient to constrain such a
model. However, we note that the image not affected by dust (image C)
in the Keck AO data is consistent with negligible shear (i.e. shear
strength $\Gamma_s=0.002$), when modeled with the pixelized technique
of \citet{V09}.

@@

%\end{minipage}
\medskip
Comparison between radio data and model-predicted flux ratios. The radio observation data is quoted from Tables 2 and Table 3 in \citet{Marlow99}. 1-SIE model is quoted from Table 5 in \citet{Marlow99}.

@@

In this letter, we focus on the effect of more complicated mass
distributions for the lens galaxy on flux ratio anomalies.
Traditionally, flux anomalous lensed quasars are radio loud systems
that are firstly discovered in radio observations. While radio data
provide reliable information on the positions and the fluxes of the
lensed images, they are often unable to detect the lens
galaxies. Moreover, lens modeling based only on the positions of the
four images can be highly degenerate and prior assumptions play a key
role in the interpretation of the observed flux ratio anomaly.  Since
lens galaxies are typically at redshifts of $z \leq 1$, optical and
near-infrared observations are ideal for measuring the light distribution
of lens galaxies and obtaining useful prior information on the underlying
mass distributions. Indeed \citet{SHARP12} showed that high
resolution IR imaging can help to significantly improve the precision
of the lens modeling. This improvement is even more dramatic in the
case of B1555+375, where HST and SHARP AO high resolution imaging has
allowed us to detect, for the fist time, the presence of an edge-on
disk. This new discovered edge-on disk provides a strong prior for our
lens modeling.  With a bulge plus an edge-on disk, our models produce
matches to both the radio image positions and their flux ratio, while
models with only one SIE component \citep{Marlow99,Xu14} fail to
reproduce the fluxes correctly. Our successful modeling indicates
that the kiloparsec-scale structure of the lens galaxy itself is
sufficient to explain the flux ratio anomalies in B1555+375, without
any need to invoke the presence of substructure.

From a theoretical perspective, \citet{Xu14} also find that substructures
cannot explain the observed radio flux anomalies in all the lenses
that were considered. After adding subhalos from numerical simulations
to these systems, only one out of the eight radio loud systems
achieves the level of the flux anomalies seen in the observations.  The rest of
them are unlikely (only $1 \% - 4 \%$ probability) to have flux
anomalies produced only by substructure, including our target B1555+375.
In fact, among those eight radio lenses, B1555+375 is one of the most
anomalous systems, but it is also one of the most challenging systems
to explain by substructures. \citet{Xu14} have concluded that the failure in
predicting flux ratios is most likely due to oversimplified/improper lens
modeling. They also point out that the SIE component and the external
shear in their model are orthogonal, indicating a possible missing
ingredient in the lens model. Their results are consistent with the SHARP
$K^{\prime}$ band observations, in which an edge-on disk crosses the
radio merging double and which fully accounts for the observed flux ratio anomaly.  In general, it is well known 
\citep[e.g,][]{Ka91} that lens models based only on the locations of lensed
quasar/AGN images have large degeneracies. Therefore, before
concluding that the flux ratios in any given lens system are a result
of substructure, it is critical to explore a range of smooth models
that are informed by additional observations or to do a comprehensive
analysis such as that conducted by \citet{Xu14}. Otherwise, derived
substructure fractions should be considered as upper limits.

@@

%. Modeling with only four image positions from radio data can also cause a lot of degeneracy. With the first system analyzing and finding an edge-on disk originated flux ratio anomaly, our follow up will be explore other similar SHARP systems. In Figure 3, all of these systems have a recognizable edge-on disk in infrared imaging. An edge-on disk structure may also be able to explain the flux ratio anomalies of these systems.
 %From the successful modeling of B1555+375,  we expect more flux anomalous systems requiring complex lens models (e.g. include an exponential disk), rather than just assuming smooth potential. Although the number of radio loud flux anomalous lensed quasars remains small, a new technique using narrow-line emissions will enlarge the sample and bring statistical perspective on substructures in the near future \citep{N14}. If the non-substructure effect from edge-on disks does not included, expected ``false'' substructure detections can serverly bias the data analysis. 
% This breaks the idea that in strongly lensed quasars, radio flux ratio
% anomalies are most likely originated from substructure
% perturbations. Our discovery shows a complex lens galaxy mass
% distribution can dominate flux ratio anomalous phenomena in a strongly
% lensed system.
% We would like to draw attention to the possible impacts on
% substructure searching from this enge-on disk orginated flux
% anomalies. It is a sign that researchers should be careful about the
% flux anomalies fraction in lensed quasars. It may be worth considering
% an upper limit of flux anomalies fraciton from observation results to
% avoid over estimate the number of substructure detections. For further
% analysis on substurcture population, mass function etc., it is always
% important to rule out other non-substructure effects.
